# Introduction

Urbanization, driven by globalization and industrialization, leads to higher population densities and significant changes in landscapes (@carnahan1974urbanization, @berry2008urbanization, @davis2015urbanization). However, urbanization also introduces anthropogenic challenges, such as light and noise pollution, which are closely tied to densely populated areas and have well-established negative effects on human, biodiversity (@sordello2020evidence) and ecosystem health. As cities expand, the need for humans to reconnect with nature becomes increasingly critical (@pyle2003nature, @buxton2024mental). This reinforces the need for accessible natural spaces within cities. In response, urban planners are integrating green spaces into cities by creating parks and preserving natural areas (@mata2020bringing). Whether urban green spaces can effectively sustain and represent native biodiversity is still being studied as urbanization evolves. Although managed natural areas and green spaces may offer critical habitats, excessive noise pollution can render these spaces unsuitable for species dependent on quiet environments for communication, reproduction, and survival. Even adaptable species face challenges when urban nature design is incomplete or poorly integrated, affecting their fitness and reproductive success.

Modern urban planning offers an opportunity to address these challenges in a holistic manner. By prioritizing quieter, functional urban spaces through district planning and zoning bylaws, urban designs can reduce noise pollution while offering residents easy access to amenities, including grocery stores, schools, and parks. Public transit improvements indirectly mitigates traffic noise, contributing to healthier urban soundscapes. To reshape urban infrastructure effectively, it is essential to understand the temporal changes and behavioural impacts of noise pollution on both humans and wildlife, which requires continuous, robust environmental monitoring. Advances in acoustic and image monitoring technologies now make such environmental assessments more feasible and cost-effective (@buxton2018pairing). However, current monitoring practices often rely on resource-intensive methods, such as moving equipment by trucks, which can contradict environmental stewardship principles.

In Alberta's 2021 Census, 82.3% of the population resided in urban areas, with a significant concentration in census metropolitan areas (CMAs) like Edmonton. The City of Edmonton has committed to enhancing its biodiversity through various strategic initiatives, including the [Natural Connections Strategic Plan, 2007](https://www.edmonton.ca/sites/default/files/public-files/assets/PDF/Natural_Connections_-_Strategic_Plan_JUNE_09.pdf?cb=1738466394) and the [Biodiversity Action Plan, 2009](https://www.edmonton.ca/sites/default/files/public-files/assets/PDF/Edmonton_Biodiversity_Action_Plan_Final.PDF). In conjunction with the tenants of these plan, this study outlines a long-term environmental monitoring program that also redefines how monitoring is conducted. By implementing zero-emission transportation and employing autonomous recording units, this program will target both noisy and non-noisy and high and low quality habitat areas of Edmonton to monitor the cities' evolving soundscape as infrastructure changes. These efforts align with the city’s broader vision for increased walking and cycling infrastructure, reducing traffic noise, and fostering sustainable practices.

This study aims to assess the proportion of natural areas in Edmonton that exceed noise sensitivity thresholds for both people and biodiversity, mapping these thresholds to guide future urban planning decisions. By shifting infrastructure investments from noise mitigation toward sustainable transit, we can create urban environments that are not only quieter but also more harmonious for both human and wildlife populations. Achieving this vision will position Edmonton as a leader in urban biodiversity preservation, sustainable urban living, and environmental stewardship.

# Methods

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| include: true
#| fig-align: center
#| fig-width: 10
#| fig-height: 10
#| fig-cap: Monitoring locations
#| label: fig-monitoring-locations

coe_2024_locs_buffered <- coe_2024_locs |> 
  sf::st_buffer(dist = 400) |> 
  sf::st_transform(crs = 4326)

coe_2025_locs_buffered <- coe_ini_2025 |> 
  sf::st_buffer(dist = 400) |> 
  sf::st_transform(crs = 4326)

leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addPolygons(
    data = coe_2025_locs_buffered,
    popup = ~paste("Location:", coe_2025_locs_buffered$location),
    color = "blue", # Buffer color
    fillOpacity = 0.3
  ) %>%
  addMeasure(primaryLengthUnit = "meters", primaryAreaUnit = "sqmeters") %>%
  addMiniMap(position = "bottomleft")
```

## Site selection and study design

The methodology for identifying potential and final monitoring sites was based on several overlapping objectives related to goals of the overall long-term monitoring study. We studied an area representing the City of Edmonton with a focus on natural areas. We wanted to ensure we selected monitoring sites captured as much of the city's noise levels variability as possible across a variety of proportion land uses related to sound. 

```{r}

# Calculated outside R for now
#dem <- st_read("geospatial_assets/DEM/DEM_Points_3TM.shp")
#dem_wgs84 <- st_transform(dem, crs = 4326)
#intersection_result <- st_intersection(dem_wgs84, coe_ini_2025)

veg_areas <- read_csv("geospatial_assets/Vegetation_Areas_-_Naturalized_20250217.csv") |>
  st_as_sf(wkt = "Multipolygon", crs = 4326)

bike_routes <- st_read("geospatial_assets/Bike_Routes/Bike_Routes.shp")

roads <- st_read("geospatial_assets/Roads/Roads.shp")

# Define custom colors: green for "Parks" and purple for others
custom_colors <- c("Parks" = "#4CAF50", "Other" = "#9C27B0") # Vibrant purple for "Other"

# Plot the spatial data with the additional points layer
ggplot(data = veg_areas) +
  geom_sf(aes(fill = ifelse(Maintainer == "Parks", "Parks", "Other")), alpha = 0.7) + # Polygons
  geom_sf(data = bike_routes, aes(color = "grey", size = 1)) +
  geom_sf(data = roads, aes(color = "black", size = 2)) +
  geom_sf(data = coe_ini_2025, aes(color = "Locations"), size = 4) + # Points layer 
  scale_fill_manual(values = custom_colors, name = "Maintainer") +
  scale_color_manual(values = c("Locations" = "blue"), name = "Locations") + # Blue color for points
  theme_minimal() +
  labs(
    title = "Parks Sites, Bike Routes and Initial Locations",
    fill = "Maintainer",
    color = "Legend"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 12)
  )


```

## Mapping noise thresholds



## Zero-emissions fieldwork

We used [Strava API](https://www.strava.com/settings/api) through [rstrava](https://github.com/fawda123/rStrava) to map the bike trips we took. We then estimated the city-miles traveled in a gas-powered F-150, a standard vehicle used for city operations. We also calculate the average road infrastructure cost savings performed instead of using vehicles.

```{r}
calculate_f150_co2 <- function(distance_km) {
  # Assumptions
  fuel_efficiency_l_per_100km <- 15.7  # Ford F-150 city fuel efficiency (L/100 km)
  co2_per_liter <- 2.31               # kg CO2 per liter for gasoline
  # Convert distance to fuel consumed (liters)
  fuel_used <- (distance_km / 100) * fuel_efficiency_l_per_100km
  # Calculate CO2 emissions (kg)
  co2_emissions <- fuel_used * co2_per_liter
  return(co2_emissions)
}

distance_km <- 100
co2_emissions <- calculate_f150_co2(distance_km)
co2_emissions

#stoken <- httr::config(token = strava_oauth("emissions", app_client_id = "147439", app_secret = Sys.getenv("STRAVA_SECRET"), app_scope = "activity:read_all"))


#stoken <- httr::config(token = strava_oauth(app_name, app_client_id, app_secret, app_scope="activity:read_all"))
#my_acts <- get_activity_list(stoken, after = as.Date('2020-12-31'))

calculate_savings <- function(daily_km, days_per_year = 250) {
  total_annual_cost <- 462.1e6  
  avg_km_per_vehicle <- 15200   
  num_vehicles <- 700000
  
  # Total annual VKT
  total_vkt <- avg_km_per_vehicle * num_vehicles  # in km
  # Road cost per km
  cost_per_km <- total_annual_cost / total_vkt  # in dollars per km
  # Annual distance biked instead of driven
  annual_biked_km <- daily_km * days_per_year  # in km
  # Annual savings
  savings <- annual_biked_km * cost_per_km  # in dollars
  return(savings)
}

# Example usage
daily_km <- 10
savings <- calculate_savings(daily_km)
print(savings)

```

# Results

Stay tuned for results as they unfold!

# Discussion

Stay tuned for results as they unfold!
