---
title: "Analyzing natural area sensitivity thresholds for humans and species diversity in Alberta’s Capital City"
format:
  html:
    grid:
      margin-width: 300px
navbar: right
theme: cosmo
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: 
  - name: "Alex MacPhail"
    affiliation:
    - "Biodiversity Pathways Ltd."
  - name: "Catherine Shier"
    affiliation:
      - "City of Edmonton"
  - name: "Erin Bayne"
    affiliation:
      - "University of Alberta, Alberta Biodiversity Monitoring Institute"
editor: visual
bibliography: references.bib
nocite: '@*'
toc: true
toc-depth: 3
toc-expand: true
toc-location: left
styles: styles.css
github: https://github.com/biodiversitypathways/coe
---

```{r}
#| label: Load packages and data
#| include: false
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(tidyverse)
library(leaflet)
library(wildrtrax)
library(sf)
library(ggridges)
library(scales)
library(kableExtra)
library(plotly)
library(DT)

load('coe.RData')
#save.image('coe.RData')

```

```{r}
#| label: Authenticate to WildTrax and download data
#| include: false
#| echo: false
#| eval: false
#| warning: false
#| message: false

wt_auth()

coe_2024 <- wt_download_report(1750, 'ARU', 'main', T)
coe_ini_2025 <- read_csv("urban_monitoring_initial_2025.csv")


```

# Abstract

# Introduction

Urbanization, driven by globalization and industrialization, leads to higher population densities and significant changes in landscapes (@carnahan1974urbanization, @berry2008urbanization, @davis2015urbanization). However, urbanization also introduces anthropogenic challenges, such as light and noise pollution, which are closely tied to densely populated areas and have well-established negative effects on human, biodiversity (@sordello2020evidence) and ecosystem health. As cities expand, the need for humans to reconnect with nature becomes increasingly critical (@pyle2003nature, @buxton2024mental). This reinforces the need for accessible natural spaces within cities. In response, urban planners are integrating green spaces into cities by creating parks and preserving natural areas (@mata2020bringing). Whether urban green spaces can effectively sustain and represent native biodiversity is still being studied as urbanization evolves. Although managed natural areas and green spaces may offer critical habitats, excessive noise pollution can render these spaces unsuitable for species dependent on quiet environments for communication, reproduction, and survival. Even adaptable species face challenges when urban nature design is incomplete or poorly integrated, affecting their fitness and reproductive success.

Modern urban planning offers an opportunity to address these challenges in a holistic manner. By prioritizing quieter, functional urban spaces through district planning and zoning bylaws, urban designs can reduce noise pollution while offering residents easy access to amenities, including grocery stores, schools, and parks. Public transit improvements indirectly mitigates traffic noise, contributing to healthier urban soundscapes. To reshape urban infrastructure effectively, it is essential to understand the temporal changes and behavioural impacts of noise pollution on both humans and wildlife, which requires continuous, robust environmental monitoring. Advances in acoustic and image monitoring technologies now make such environmental assessments more feasible and cost-effective (@buxton2018pairing). However, current monitoring practices often rely on resource-intensive methods, such as moving equipment by trucks, which can contradict environmental stewardship principles.

In Alberta's 2021 Census, 82.3% of the population resided in urban areas, with a significant concentration in census metropolitan areas (CMAs) like Edmonton. The City of Edmonton has committed to enhancing its biodiversity through various strategic initiatives, including the [Natural Connections Strategic Plan, 2007](https://www.edmonton.ca/sites/default/files/public-files/assets/PDF/Natural_Connections_-_Strategic_Plan_JUNE_09.pdf?cb=1738466394) and the [Biodiversity Action Plan, 2009](https://www.edmonton.ca/sites/default/files/public-files/assets/PDF/Edmonton_Biodiversity_Action_Plan_Final.PDF). In conjunction with the tenants of these plan, this study outlines a long-term environmental monitoring program that also redefines how monitoring is conducted. By implementing zero-emission transportation and employing autonomous recording units, this program will target both noisy and non-noisy and high and low quality habitat areas of Edmonton to monitor the cities' evolving soundscape as infrastructure changes. These efforts align with the city’s broader vision for increased walking and cycling infrastructure, reducing traffic noise, and fostering sustainable practices.

This study aims to assess the proportion of natural areas in Edmonton that exceed noise sensitivity thresholds for both people and biodiversity, mapping these thresholds to guide future urban planning decisions. By shifting infrastructure investments from noise mitigation toward sustainable transit, we can create urban environments that are not only quieter but also more harmonious for both human and wildlife populations. Achieving this vision will position Edmonton as a leader in urban biodiversity preservation, sustainable urban living, and environmental stewardship.

# Methods

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| include: true
#| fig-align: center
#| fig-width: 10
#| fig-height: 10
#| fig-cap: Monitoring locations
#| label: fig-monitoring-locations


coe_2024_locs <- coe_2024 |>
  select(location, latitude, longitude) |>
  distinct() |>
  sf::st_as_sf(coords = c("longitude","latitude"), crs = 4326)

coe_ini_2025 <- coe_ini_2025 %>%
  st_as_sf(wkt = "WKT", crs = 4326)

leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addCircleMarkers(
    data = coe_2024_locs,
    popup = ~paste("Location:", coe_2024_locs$location),
    radius = 5,
    color = "blue"  # Correct argument name
  ) %>%
  addCircleMarkers(
    data = coe_ini_2025,
    popup = ~paste("Location:", coe_ini_2025$location),
    radius = 5,
    color = "green"  # Correct argument name
  ) %>%
  addMeasure(primaryLengthUnit = "meters", primaryAreaUnit = "sqmeters") %>%
  addMiniMap(position = "bottomleft")


```

## Site selection and study design

The methodology for identifying potential and final monitoring sites was based on several overlapping objectives related to goals of the overall study. First, natural areas were the primarily selection criteria. Using several GIS approaches were combined to satisfy these objectives: Multi-criteria analysis (MCA), location-allocation modelling (LAM), and constrained random allocation, natural areas were categorized into a categorical gradient of noise conditions. This entailed ensuring monitoring sites captured as much of the city as possible, noise levels where residents are concentrated, and a variety of land uses. Complementing this were categorical candidate sites of particular and general interest. 

```{r}

dem <- st_read("./DEM/DEM_Points_3TM.shp")

veg_areas <- read_csv("Vegetation_Areas_-_Naturalized_20250217.csv") |>
  st_as_sf(wkt = "Multipolygon", crs = 4326)

# Define custom colors: green for "Parks" and purple for others
custom_colors <- c("Parks" = "#4CAF50", "Other" = "#9C27B0") # Vibrant purple for "Other"

# Plot the spatial data with the additional points layer
ggplot(data = veg_areas) +
  geom_sf(aes(fill = ifelse(Maintainer == "Parks", "Parks", "Other")), color = "black", alpha = 0.7) + # Polygons
  geom_sf(data = coe_ini_2025, aes(color = "Locations"), size = 2) + # Points layer
  scale_fill_manual(values = custom_colors, name = "Maintainer") +
  scale_color_manual(values = c("Locations" = "blue"), name = "Locations") + # Blue color for points
  theme_minimal() +
  labs(
    title = "Vegetation Areas with Locations",
    fill = "Maintainer",
    color = "Legend"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 12)
  )


```


### Zero-emissions fieldwork

```{r}
calculate_f150_co2 <- function(distance_km) {
  # Assumptions
  fuel_efficiency_l_per_100km <- 15.7  # Ford F-150 city fuel efficiency (L/100 km)
  co2_per_liter <- 2.31               # kg CO2 per liter for gasoline
  # Convert distance to fuel consumed (liters)
  fuel_used <- (distance_km / 100) * fuel_efficiency_l_per_100km
  # Calculate CO2 emissions (kg)
  co2_emissions <- fuel_used * co2_per_liter
  return(co2_emissions)
}

distance_km <- 100
co2_emissions <- calculate_f150_co2(distance_km)
co2_emissions

stoken <- httr::config(token = strava_oauth("emissions", app_client_id = "147439", app_secret = Sys.getenv("STRAVA_SECRET"), app_scope = "activity:read_all"))


stoken <- httr::config(token = strava_oauth(app_name, app_client_id, app_secret, app_scope="activity:read_all"))
my_acts <- get_activity_list(stoken, after = as.Date('2020-12-31'))


```

## Mapping noise thresholds

# Results

# Discussion
